---
title: "Income"
---

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
library(tidyverse)
reticulate::py_require(c("numpy", "polars"))
```

```{python}
#| label: setup-py
#| include: false
import polars as pl
import numpy as np
```

## Math for Income {#math-for-income}

The [income](https://mjfrigaard.github.io/fm-finance/income.html) chapter is mostly about behavior, but it also introduces some math and how to write R/Python functions.

### Functions {.unnumbered}

Functions perform operations: calculate a total, build a table, create a graph, etc. The function's *arguments* are the inputs it needs to do its job, and the *output* is what it gives back. For example, the `sum()` function takes a list of numbers as an argument and outputs their total.

In R, functions are defined using the `function()` keyword,

``` r
my_fun <- function() {
  # function body
}
```

In Python, you use `def`:

``` python
def my_fun():
    # function body
```

Both languages allow you to create your own functions to perform specific tasks, which is especially useful for financial calculations.

### The Fixed-Cost Ratio {.unnumbered}

Fixed costs are things like rent, loan payments, insurance, and utilities. THey are considered *fixed* because they don't really change month to month.

A good benchmark is to keep them under \~60% of your income/take-home pay (leaving about 40% for *variable* costs and savings).

**Formula:** Fixed-Cost Ratio = Fixed Costs ÷ Take-Home Pay × 100

Example: \$2,700 of fixed costs on \$5,000 take-home → 54%, just under the ceiling.

::: panel-tabset

## R

```{r}
#| label: fixed-cost-r
fixed_cost_ratio <- function(fixed_costs, take_home) {
  (fixed_costs / take_home) * 100
}
```

Now we can apply the function to \$2,700 of fixed costs on \$5,000 take-home pay. The output is a percentage, so we can add a `%` sign to the output with the `paste0()` function:

```{r}
#| label: fixed-cost-r-apply
fixed_cost_ratio(fixed_costs = 2700, take_home = 5000) |> 
  paste0("%")
```

#### Using pipes in R {.unnumbered}

The pipe operator (`|>`) allows you to take the output of one function and use it as the input for another function, making your code more readable and efficient. Read more in the [Pipe Operator](r_vs_py.qmd#sec-pipe-operator) section of R vs. Python.

``` r
function1() |> 
  function2() |> 
  function3()
```

In the example above, we calculate the fixed-cost ratio and then immediately format it as a percentage string. A better way to do this is to create a new function that combines both steps:

```{r}
#| label: fixed-cost-r-pipe
fixed_cost_ratio_pct <- function(fixed_costs, take_home) {
  ratio <- (fixed_costs / take_home) * 100
  paste0(ratio, "%")
}
fixed_cost_ratio_pct(fixed_costs = 2700, take_home = 5000)
```

## Python

In Python, we can define the same function without needing a pipe operator, since we can directly return the formatted string:

```{python}
#| label: fixed-cost-py
def fixed_cost_ratio(fixed_costs, take_home):
    ratio = (fixed_costs / take_home) * 100
    return f"{ratio:.2f}%"
```

\$2,700 of fixed costs on \$5,000 take-home pay

```{python}
#| label: fixed-cost-py-apply
fixed_cost_ratio(fixed_costs=2700, take_home=5000)
```

Note the differences in using `paste0()` in R vs. f-strings in Python for formatting the output as a percentage string. Read more about string formatting in [String Formatting](r_vs_py.qmd#sec-string-formatting).

## Excel

In Excel there's no function to define: you put the numbers in cells and write the formula once. Enter the fixed costs in cell **A2** and take-home pay in **B2**, then in **C2**:

```excel
=A2/B2
```

That gives `0.54`. Format **C2** as a percentage (**Home → %**, or **Ctrl+Shift+5**) and it displays **54%**.

To produce the percent *string* directly (the analog of R's `paste0(..., "%")` and Python's f-string), use `TEXT()` instead:

```excel
=TEXT(A2/B2, "0.00%")
```

This returns the text `"54.00%"`, matching the Python output above.

:::

### What a Raise Is Really Worth {.unnumbered}

In the [Income](https://mjfrigaard.github.io/fm-finance/income.html) chapter, we saw that "Split the raise" (i.e., put half of your monthly raise into the market for a few decades) is a powerful strategy for building wealth. But how powerful? Let's put some numbers to it.

**Formula:** FV = PMT × \[((1 + r)\^n − 1) ÷ r\]

- PMT is the amount invested each month  
- *r* is the monthly return  
- *n* is the number of months.

This is the future-value-of-contributions formula we revisit in [Investing Basics](https://mjfrigaard.github.io/fm-finance/investing_basics.html). 

::: panel-tabset

## R

```{r}
#| label: raise-value-r
raise_invested_value <- function(monthly_raise, invest_share, rate, years) {
  contribution <- monthly_raise * invest_share
  monthly_rate <- rate / 12
  periods      <- years * 12
  contribution * (((1 + monthly_rate)^periods - 1) / monthly_rate)
}
```

Let's see what happens when we take half of a $300/month raise, invested for 15 years at 7%:

```{r}
#| label: raise-value-r-apply
raise_invested_value(monthly_raise = 300, invest_share = 0.5, rate = 0.07, years = 15)
```

## Python

```{python}
#| label: raise-value-py
def raise_invested_value(monthly_raise, invest_share, rate, years):
    contribution = monthly_raise * invest_share
    monthly_rate = rate / 12
    periods = years * 12
    return contribution * (((1 + monthly_rate) ** periods - 1) / monthly_rate)
```

Half of a $300/month raise, invested for 15 years at 7%

```{python}
#| label: raise-value-py-apply
raise_invested_value(monthly_raise=300, invest_share=0.5, rate=0.07, years=15)
```

## Excel

This is exactly what Excel's built-in **`FV()`** (future value) function is for, so you don't have to write the formula yourself. Its arguments are `FV(rate, nper, pmt)`:

- `rate` --- the return *per period* (monthly here): `7%/12`
- `nper` --- the number of periods: `25*12`
- `pmt` --- the amount invested each period, entered as a **negative** number because it's cash leaving your pocket: `-200`

```excel
=FV(7%/12, 25*12, -200)
```

This returns **\$162,014.34**, the same result as the R and Python functions above. (Excel reports a positive number because the contributions were entered as negative cash flows.)

:::