---
title: "Income"
---

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
library(tidyverse)
reticulate::py_require(c("numpy", "polars"))
```

```{python}
#| label: setup-py
#| include: false
import polars as pl
import numpy as np
```

## Math for Income {#math-for-income}

The income chapter is mostly about behavior, but two numbers turn that behavior into something you can track. Each calculation below is also written as an R and Python function, following the same pattern introduced in the [Budgeting](budgeting.qmd) chapter.

### The Fixed-Cost Ratio {.unnumbered}

Fixed costs (rent, loan payments, insurance, utilities) are the commitments you can't easily change month to month. Keeping them under \~60% of take-home pay leaves room to save *and* to spend on what you love without guilt.

**Formula:** Fixed-Cost Ratio = Fixed Costs ÷ Take-Home Pay × 100

Example: \$2,700 of fixed costs on \$5,000 take-home → 54%, just under the ceiling.

::: panel-tabset
## R

```{r}
#| label: fixed-cost-r
fixed_cost_ratio <- function(fixed_costs, take_home) {
  (fixed_costs / take_home) * 100
}

# $2,700 of fixed costs on $5,000 take-home pay
fixed_cost_ratio(fixed_costs = 2700, take_home = 5000)
```

## Python

```{python}
#| label: fixed-cost-py
def fixed_cost_ratio(fixed_costs, take_home):
    return (fixed_costs / take_home) * 100

# $2,700 of fixed costs on $5,000 take-home pay
fixed_cost_ratio(fixed_costs=2700, take_home=5000)
```
:::

### What a Raise Is Really Worth {.unnumbered}

"Split the raise" only pays off if the saved half is actually invested. Putting half of a monthly raise into the market for a few decades shows why the *habit* matters more than the dollar amount.

**Formula:** FV = PMT × [((1 + r)\^n − 1) ÷ r]

This is the future-value-of-contributions formula from [Investing Basics](investing_basics.qmd), where PMT is the amount invested each month, *r* is the monthly return, and *n* is the number of months.

Example: a \$400/month raise, half of it invested (\$200/month) for 25 years at a 7% annual return → **\~\$162,000**.

::: panel-tabset
## R

```{r}
#| label: raise-value-r
raise_invested_value <- function(monthly_raise, invest_share, rate, years) {
  contribution <- monthly_raise * invest_share
  monthly_rate <- rate / 12
  periods      <- years * 12
  contribution * (((1 + monthly_rate)^periods - 1) / monthly_rate)
}

# half of a $400/month raise, invested for 25 years at 7%
raise_invested_value(monthly_raise = 400, invest_share = 0.5, rate = 0.07, years = 25)
```

## Python

```{python}
#| label: raise-value-py
def raise_invested_value(monthly_raise, invest_share, rate, years):
    contribution = monthly_raise * invest_share
    monthly_rate = rate / 12
    periods = years * 12
    return contribution * (((1 + monthly_rate) ** periods - 1) / monthly_rate)

# half of a $400/month raise, invested for 25 years at 7%
raise_invested_value(monthly_raise=400, invest_share=0.5, rate=0.07, years=25)
```
:::

