---
title: "Investing Basics"
---

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

## Book Topics

- **Compounding**: interest earning interest; does the heavy lifting over long horizons; Rule of 72 — years to double ≈ 72 ÷ annual return %
- **Risk vs. return**: higher expected return = higher volatility; choose a level you can *stick with* through downturns
- **Diversification**: don't bet on one company, sector, or country; broad index funds make diversification nearly free (Bogle)
- **Real vs. nominal return**: inflation silently erodes purchasing power; always think in real (inflation-adjusted) terms

## Future Value (Compounding)

A single lump sum growing at a constant rate.

**Formula:** FV = PV × (1 + r)\^n

```{r}
#| label: compound-bar-model
#| echo: false
#| fig-cap: "$10,000 growing at 7% — value at each checkpoint"
#| fig-asp: 0.50
library(ggplot2)
yrs  <- c(0, 5, 10, 20, 30)
vals <- round(10000 * (1.07)^yrs)
bar_colors <- c("#86ddcd", "#5ec8be", "#0e9aa7", "#e07070", "#f44242")
df_fv <- data.frame(
  year  = factor(paste0("Year ", yrs), levels = paste0("Year ", yrs)),
  value = vals,
  fill  = bar_colors
)
ggplot(df_fv, aes(x = year, y = value, fill = fill)) +
  geom_col(width = 0.6, colour = "#1C1C1E", linewidth = 0.4) +
  geom_text(aes(label = paste0("$", formatC(value, format = "d", big.mark = ","))),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  scale_fill_identity() +
  scale_y_continuous(limits = c(0, max(vals) * 1.18),
                     labels = function(x) paste0("$", round(x / 1000), "k")) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor = element_blank())
```

::: panel-tabset
## R

```{r}
#| label: compound-r
future_value <- function(present_value, rate, years) {
  present_value * (1 + rate)^years
}

rule_of_72 <- function(annual_return_pct) {
  72 / annual_return_pct
}

future_value(present_value = 10000, rate = 0.07, years = 10)
rule_of_72(annual_return_pct = 7)

# vectorized across multiple horizons
data.frame(
  years = c(5, 10, 20, 30),
  value = future_value(10000, 0.07, c(5, 10, 20, 30))
)
```

## Python

```{python}
#| label: compound-py
import numpy as np

def future_value(present_value, rate, years):
    return present_value * (1 + rate) ** years

def rule_of_72(annual_return_pct):
    return 72 / annual_return_pct

print(future_value(present_value=10000, rate=0.07, years=10))
print(rule_of_72(annual_return_pct=7))

horizons = np.array([5, 10, 20, 30])
future_value(present_value=10000, rate=0.07, years=horizons)
```

## Excel

With present value in `A2`, rate in `B2`, years in `C2`:

```swift
=A2*(1+B2)^C2
```

Rule of 72 (rate as a percent in `A2`):

```swift
=72/A2
```
:::

## Future Value of Regular Contributions

Monthly investing builds wealth two ways: the money contributed and the returns on all prior contributions.

**Formula:** FV = PMT × \[((1 + r)\^n − 1) ÷ r\]

```{r}
#| label: fv-series-bar-model
#| echo: false
#| fig-cap: "$500/month contributions at 7% — contributed vs. gains"
#| fig-asp: 0.50
library(ggplot2)
yrs_vec    <- c(10, 20, 30)
annual_pmt <- 500 * 12
contributed <- annual_pmt * yrs_vec
total       <- annual_pmt * ((1.07^yrs_vec - 1) / 0.07)
gains       <- total - contributed
df_long <- data.frame(
  year   = rep(factor(paste0("Year ", yrs_vec), levels = paste0("Year ", yrs_vec)), 2),
  amount = c(contributed, gains),
  layer  = factor(rep(c("Money contributed", "Investment gains"), each = 3),
                  levels = c("Investment gains", "Money contributed"))
)
totals_lab <- data.frame(
  year  = factor(paste0("Year ", yrs_vec), levels = paste0("Year ", yrs_vec)),
  total = contributed + gains
)
ggplot(df_long, aes(x = year, y = amount, fill = layer)) +
  geom_col(width = 0.6, colour = "#1C1C1E", linewidth = 0.4) +
  geom_text(data = totals_lab, aes(x = year, y = total,
                label = paste0("$", formatC(round(total), format = "d", big.mark = ","))),
            inherit.aes = FALSE, vjust = -0.4, fontface = "bold", size = 3.5) +
  scale_fill_manual(values = c("Money contributed" = "#0e9aa7",
                               "Investment gains"  = "#86ddcd")) +
  scale_y_continuous(limits = c(0, max(contributed + gains) * 1.15),
                     labels = function(x) paste0("$", round(x / 1000), "k")) +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "top")
```

::: panel-tabset
## R

```{r}
#| label: fv-series-r
future_value_series <- function(contribution, rate, periods) {
  contribution * (((1 + rate)^periods - 1) / rate)
}

# $500/month for 30 years at 7% annual (monthly rate = 0.07/12)
future_value_series(contribution = 500, rate = 0.07 / 12, periods = 30 * 12)
```

## Python

```{python}
#| label: fv-series-py
def future_value_series(contribution, rate, periods):
    return contribution * (((1 + rate) ** periods - 1) / rate)

future_value_series(contribution=500, rate=0.07 / 12, periods=30 * 12)
```

## Excel

Use the built-in `FV()` function: `FV(rate, nper, pmt)`

- `rate`: monthly = `7%/12`
- `nper`: months = `30*12`
- `pmt`: monthly contribution as negative = `-500`

```swift
=FV(7%/12, 30*12, -500)
```
:::

## Real (Inflation-Adjusted) Return

Nominal returns overstate purchasing power gains; inflation takes its slice every year.

**Formula:** Real Return = (1 + nominal) ÷ (1 + inflation) − 1

```{r}
#| label: real-return-bar-model
#| echo: false
#| fig-cap: "Nominal return, inflation, and real return"
#| fig-asp: 0.50
library(ggplot2)
df_rr <- data.frame(
  label = factor(c("Nominal return (7%)", "Inflation (3%)", "Real return (~3.9%)"),
                 levels = c("Nominal return (7%)", "Inflation (3%)", "Real return (~3.9%)")),
  value = c(7, 3, (1.07 / 1.03 - 1) * 100),
  fill  = c("#0e9aa7", "#f44242", "#48a56a")
)
ggplot(df_rr, aes(x = label, y = value, fill = fill)) +
  geom_col(width = 0.55, colour = "#1C1C1E", linewidth = 0.4) +
  geom_text(aes(label = paste0(round(value, 1), "%")),
            vjust = -0.4, fontface = "bold", size = 4) +
  scale_fill_identity() +
  scale_y_continuous(limits = c(0, 9), labels = function(x) paste0(x, "%")) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor = element_blank())
```

::: panel-tabset
## R

```{r}
#| label: real-return-r
real_return <- function(nominal, inflation) {
  (1 + nominal) / (1 + inflation) - 1
}

real_return(nominal = 0.07, inflation = 0.03)
```

## Python

```{python}
#| label: real-return-py
def real_return(nominal, inflation):
    return (1 + nominal) / (1 + inflation) - 1

real_return(nominal=0.07, inflation=0.03)
```

## Excel

With nominal rate in `A2` and inflation in `B2`:

```swift
=(1+A2)/(1+B2)-1
```
:::
