---
title: "Investing Basics"
---

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
library(tidyverse)
reticulate::py_require(c("numpy", "polars"))
```

```{python}
#| label: setup-py
#| include: false
import polars as pl
import numpy as np
```

## Math for Investing Basics {#math-for-investing-basics}

Investing math is mostly one idea, compounding, applied in a few different ways. Each calculation below is written as an R and Python function, following the same pattern introduced in the [Budgeting](budgeting.qmd) chapter.

In code, the `^` symbol (R) and `**` symbol (Python) both mean "raise to the power of." So `(1 + 0.07)^10` means "1.07 multiplied by itself 10 times."

### Future Value and the Rule of 72 {.unnumbered}

**Formula:** FV = PV × (1 + r)\^n

Where: FV = future value, PV = present value (today's amount), *r* = annual return, *n* = years.

The companion shortcut is the **Rule of 72**: divide 72 by the return percentage to estimate how many years it takes your money to double.

> Years to double ≈ 72 ÷ annual return %

Example: \$10,000 at 7% for 10 years grows to \~\$19,672, and at 7% money doubles roughly every 10 years.

::: panel-tabset
## R

```{r}
#| label: compound-r
future_value <- function(present_value, rate, years) {
  present_value * (1 + rate)^years
}

rule_of_72 <- function(annual_return_pct) {
  72 / annual_return_pct
}

# scalar: $10,000 at 7% for 10 years
future_value(present_value = 10000, rate = 0.07, years = 10)

# how long until your money doubles at 7%?
rule_of_72(annual_return_pct = 7)

# vectorized: same $10,000 at 7% across multiple horizons
data.frame(
  years = c(5, 10, 20, 30),
  value = future_value(present_value = 10000, rate = 0.07, years = c(5, 10, 20, 30))
)
```

## Python

```{python}
#| label: compound-py
def future_value(present_value, rate, years):
    return present_value * (1 + rate) ** years

def rule_of_72(annual_return_pct):
    return 72 / annual_return_pct

# scalar: $10,000 at 7% for 10 years
print(future_value(present_value=10000, rate=0.07, years=10))

# how long until your money doubles at 7%?
print(rule_of_72(annual_return_pct=7))

# vectorized: numpy lets us pass an array of horizons in one call
horizons = np.array([5, 10, 20, 30])
future_value(present_value=10000, rate=0.07, years=horizons)
```
:::

### Future Value of Regular Contributions {.unnumbered}

Most people don't invest a lump sum once; they add money every month. The future value of a stream of equal contributions is:

**Formula:** FV = PMT × [((1 + r)\^n − 1) ÷ r]

Where: PMT = the amount contributed each period, *r* = the return per period, *n* = the number of periods.

Example: \$500/month for 30 years (360 months) at a 7% annual return (\~0.583% per month) grows to \~\$610,000, of which only \$180,000 is money you actually put in.

::: panel-tabset
## R

```{r}
#| label: fv-series-r
future_value_series <- function(contribution, rate, periods) {
  contribution * (((1 + rate)^periods - 1) / rate)
}

# $500/month for 30 years at a 7% annual return (monthly rate = 0.07 / 12)
future_value_series(contribution = 500, rate = 0.07 / 12, periods = 30 * 12)
```

## Python

```{python}
#| label: fv-series-py
def future_value_series(contribution, rate, periods):
    return contribution * (((1 + rate) ** periods - 1) / rate)

# $500/month for 30 years at a 7% annual return (monthly rate = 0.07 / 12)
future_value_series(contribution=500, rate=0.07 / 12, periods=30 * 12)
```
:::

### Real (Inflation-Adjusted) Return {.unnumbered}

A 7% return with 3% inflation does *not* leave you 4% richer in spending power; the relationship is multiplicative, not subtractive.

**Formula:** Real Return = (1 + nominal) ÷ (1 + inflation) − 1

Example: (1.07 ÷ 1.03) − 1 = \~3.88%, slightly less than the 4% you'd get by simple subtraction.

::: panel-tabset
## R

```{r}
#| label: real-return-r
real_return <- function(nominal, inflation) {
  (1 + nominal) / (1 + inflation) - 1
}

# 7% nominal return against 3% inflation
real_return(nominal = 0.07, inflation = 0.03)
```

## Python

```{python}
#| label: real-return-py
def real_return(nominal, inflation):
    return (1 + nominal) / (1 + inflation) - 1

# 7% nominal return against 3% inflation
real_return(nominal=0.07, inflation=0.03)
```
:::
