---
title: "Major Purchases"
---

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
library(tidyverse)
reticulate::py_require(c("numpy", "polars"))
```

```{python}
#| label: setup-py
#| include: false
import polars as pl
import numpy as np
```

## Math for Major Purchases {#math-for-major-purchases}

Big purchases are where small percentage differences turn into tens of thousands of dollars. Each calculation below is written as an R and Python function, following the same pattern introduced in the [Budgeting](budgeting.qmd) chapter.

### The Rent-vs-Buy 5% Rule {.unnumbered}

Felix's rule estimates the annual *unrecoverable* cost of owning as about 5% of the home's value, then compares the monthly equivalent to your rent.

**Formula:** Monthly cost of owning ≈ (Home Value × 0.05) ÷ 12

Example: a \$600,000 home → (600,000 × 0.05) ÷ 12 = **\$2,500/month**. Rent below that, and renting-and-investing-the-difference tends to win.

::: panel-tabset
## R

```{r}
#| label: rent-buy-r
rent_buy_threshold <- function(home_value, unrecoverable_rate = 0.05) {
  (home_value * unrecoverable_rate) / 12
}

# $600,000 home at the standard 5% unrecoverable rate
rent_buy_threshold(home_value = 600000)
```

## Python

```{python}
#| label: rent-buy-py
def rent_buy_threshold(home_value, unrecoverable_rate=0.05):
    return (home_value * unrecoverable_rate) / 12

# $600,000 home at the standard 5% unrecoverable rate
rent_buy_threshold(home_value=600000)
```
:::

### The Monthly Mortgage Payment {.unnumbered}

This is the standard amortization formula behind every mortgage calculator.

**Formula:** M = P × [ r(1 + r)\^n ] ÷ [ (1 + r)\^n − 1 ]

Where: P = the loan principal, *r* = the monthly interest rate (annual ÷ 12), *n* = the total number of payments (years × 12).

Example: a \$480,000 loan (a \$600,000 home with 20% down) at 6% for 30 years → **\~\$2,878/month**.

::: panel-tabset
## R

```{r}
#| label: mortgage-payment-r
mortgage_payment <- function(principal, annual_rate, years) {
  r <- annual_rate / 12
  n <- years * 12
  principal * (r * (1 + r)^n) / ((1 + r)^n - 1)
}

# $480,000 loan at 6% over 30 years
mortgage_payment(principal = 480000, annual_rate = 0.06, years = 30)
```

## Python

```{python}
#| label: mortgage-payment-py
def mortgage_payment(principal, annual_rate, years):
    r = annual_rate / 12
    n = years * 12
    return principal * (r * (1 + r) ** n) / ((1 + r) ** n - 1)

# $480,000 loan at 6% over 30 years
mortgage_payment(principal=480000, annual_rate=0.06, years=30)
```
:::

### Where the First Year's Payments Go {.unnumbered}

The front-loaded-interest reality, in numbers: in month one your interest is simply the balance times the monthly rate, and only the rest reduces the principal.

**Formula:**

- First-month interest = Principal × (annual rate ÷ 12)
- First-month principal = Monthly Payment − First-month interest

Example: on that \$480,000 loan, month-one interest is \$2,400 of the \~\$2,878 payment, about **83% interest**, leaving only \~\$478 toward principal.

::: panel-tabset
## R

```{r}
#| label: first-payment-r
first_payment_split <- function(principal, annual_rate, years) {
  payment  <- mortgage_payment(principal, annual_rate, years)
  interest <- principal * (annual_rate / 12)
  data.frame(
    payment      = payment,
    interest     = interest,
    principal    = payment - interest,
    pct_interest = (interest / payment) * 100
  )
}

# first payment on a $480,000 loan at 6% over 30 years
first_payment_split(principal = 480000, annual_rate = 0.06, years = 30)
```

## Python

```{python}
#| label: first-payment-py
def first_payment_split(principal, annual_rate, years):
    payment = mortgage_payment(principal, annual_rate, years)
    interest = principal * (annual_rate / 12)
    return {
        "payment": payment,
        "interest": interest,
        "principal": payment - interest,
        "pct_interest": (interest / payment) * 100,
    }

# first payment on a $480,000 loan at 6% over 30 years
first_payment_split(principal=480000, annual_rate=0.06, years=30)
```
:::
