---
title: "Major Purchases"
---

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

## Book Topics

- **Renting vs. buying**: "renting is throwing money away" is a myth; unrecoverable costs of owning run ~5% of home value per year (Felix's 5% Rule)
- **5% Rule components**: property tax (~1%) + maintenance (~1%) + cost of capital (~3%) = ~5%/year; divide by 12 to get monthly equivalent
- **When to buy**: plan to stay 7+ years; stable income; won't be "house poor" (housing <28–30% of gross income)
- **Mortgage amortization**: early payments are mostly interest (~75–80% in year 1 on a 30-year loan); moving every 5 years means you've barely touched principal
- **Cars**: depreciating assets; 20/4/10 rule (20% down, ≤4-year term, ≤10% gross income on transportation); buy used 2–4 years old

## Rent vs. Buy Threshold (Felix's 5% Rule)

Multiply home value by 5%, divide by 12. If you can rent an equivalent home for less, renting and investing the difference is likely the better financial choice.

**Formula:** Monthly ownership cost ≈ (Home Value × 0.05) ÷ 12

```{r}
#| label: rent-buy-bar-model
#| echo: false
#| fig-cap: "The three components of the 5% rule on a $500,000 home"
#| fig-asp: 0.50
library(ggplot2)
df_5pct <- data.frame(
  component = factor(
    c("Property Tax (~1%)\n~$5,000/yr", "Maintenance (~1%)\n~$5,000/yr",
      "Cost of Capital (~3%)\n~$15,000/yr"),
    levels = c("Cost of Capital (~3%)\n~$15,000/yr",
               "Maintenance (~1%)\n~$5,000/yr",
               "Property Tax (~1%)\n~$5,000/yr")
  ),
  amount = c(5000, 5000, 15000)
)
ggplot(df_5pct, aes(x = "", y = amount, fill = component)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = component), position = position_stack(vjust = 0.5),
            colour = "#F5F2E8", fontface = "bold", size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c("Property Tax (~1%)\n~$5,000/yr" = "#f44242",
                               "Maintenance (~1%)\n~$5,000/yr"  = "#0e9aa7",
                               "Cost of Capital (~3%)\n~$15,000/yr" = "#8b1a1a")) +
  scale_y_continuous(breaks = c(0, 5000, 10000, 15000, 20000, 25000),
                     labels = function(x) paste0("$", format(x, big.mark = ","))) +
  labs(x = NULL, y = "Annual unrecoverable cost (Total = $25,000/yr = $2,083/mo)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", axis.text.y = element_blank(),
        axis.ticks.y = element_blank(), panel.grid.minor = element_blank())
```

::: panel-tabset
## R

```{r}
#| label: rent-buy-r
rent_buy_threshold <- function(home_value, unrecoverable_rate = 0.05) {
  (home_value * unrecoverable_rate) / 12
}

rent_buy_threshold(home_value = 600000)
```

## Python

```{python}
#| label: rent-buy-py
def rent_buy_threshold(home_value, unrecoverable_rate=0.05):
    return (home_value * unrecoverable_rate) / 12

rent_buy_threshold(home_value=600000)
```

## Excel

With home value in `A2`:

```swift
=(A2*0.05)/12
```
:::

## Monthly Mortgage Payment

The standard amortization formula; shorter terms mean higher payments but much less total interest.

**Formula:** M = P × \[r(1 + r)\^n\] ÷ \[(1 + r)\^n − 1\]

```{r}
#| label: mortgage-payment-bar-model
#| echo: false
#| fig-cap: "Monthly payment on a $400,000 loan at 6.5% across three terms"
#| fig-asp: 0.50
library(ggplot2)
P <- 400000; r <- 0.065 / 12; terms <- c(15, 20, 30); n_vals <- terms * 12
payments <- round(P * (r * (1 + r)^n_vals) / ((1 + r)^n_vals - 1))
df_mort <- data.frame(term = factor(paste0(terms, "-Year"), levels = paste0(terms, "-Year")),
                      payment = payments)
ggplot(df_mort, aes(x = term, y = payment, fill = term)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = paste0("$", format(payment, big.mark = ","))),
            vjust = -0.4, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("15-Year" = "#f44242", "20-Year" = "#0e9aa7",
                               "30-Year" = "#48a56a")) +
  scale_y_continuous(limits = c(0, max(payments) * 1.18),
                     labels = function(x) paste0("$", format(x, big.mark = ","))) +
  labs(x = NULL, y = "Monthly payment ($400,000 loan at 6.5%)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank())
```

::: panel-tabset
## R

```{r}
#| label: mortgage-payment-r
mortgage_payment <- function(principal, annual_rate, years) {
  r <- annual_rate / 12
  n <- years * 12
  principal * (r * (1 + r)^n) / ((1 + r)^n - 1)
}

# $480,000 loan at 6% over 30 years
mortgage_payment(principal = 480000, annual_rate = 0.06, years = 30)
```

## Python

```{python}
#| label: mortgage-payment-py
def mortgage_payment(principal, annual_rate, years):
    r = annual_rate / 12
    n = years * 12
    return principal * (r * (1 + r) ** n) / ((1 + r) ** n - 1)

mortgage_payment(principal=480000, annual_rate=0.06, years=30)
```

## Excel

Use the built-in `PMT()` function: `PMT(rate, nper, pv)`

- `rate`: monthly = `6%/12`
- `nper`: months = `30*12`
- `pv`: loan amount as negative = `-480000`

```swift
=PMT(6%/12, 30*12, -480000)
```
:::

## First-Month Interest vs. Principal Split

Early mortgage payments are mostly interest; only the remainder reduces the balance.

**Formulas:**

- First-month interest = Principal × (annual rate ÷ 12)
- First-month principal = Monthly Payment − First-month interest

```{r}
#| label: first-payment-bar-model
#| echo: false
#| fig-cap: "Interest vs. principal split at three points on a $400,000 / 6.5% / 30-yr mortgage"
#| fig-asp: 0.50
library(ggplot2)
df_amort <- data.frame(
  month  = factor(c("Month 1\n(Year 1)", "Month 180\n(Year 15)", "Month 348\n(Year 29)"),
                  levels = c("Month 1\n(Year 1)", "Month 180\n(Year 15)", "Month 348\n(Year 29)")),
  type   = rep(c("Interest", "Principal"), times = 3),
  amount = c(2167, 361, 1380, 1148, 82, 2446),
  label  = c("$2,167\n(86%)", "$361\n(14%)", "$1,380\n(55%)",
             "$1,148\n(45%)", "$82\n(3%)", "$2,446\n(97%)")
)
ggplot(df_amort, aes(x = month, y = amount, fill = type)) +
  geom_col(width = 0.55, position = "stack") +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5),
            colour = "#F5F2E8", fontface = "bold", size = 3.5) +
  scale_fill_manual(values = c("Interest" = "#f44242", "Principal" = "#48a56a")) +
  scale_y_continuous(limits = c(0, 2900),
                     labels = function(x) paste0("$", format(x, big.mark = ","))) +
  labs(x = NULL, y = "Monthly payment (~$2,528)", fill = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank())
```

::: panel-tabset
## R

```{r}
#| label: first-payment-r
first_payment_split <- function(principal, annual_rate, years) {
  payment  <- mortgage_payment(principal, annual_rate, years)
  interest <- principal * (annual_rate / 12)
  data.frame(
    payment      = round(payment, 2),
    interest     = round(interest, 2),
    principal    = round(payment - interest, 2),
    pct_interest = round((interest / payment) * 100, 1)
  )
}

first_payment_split(principal = 480000, annual_rate = 0.06, years = 30)
```

## Python

```{python}
#| label: first-payment-py
def first_payment_split(principal, annual_rate, years):
    payment = mortgage_payment(principal, annual_rate, years)
    interest = principal * (annual_rate / 12)
    return {
        "payment":      round(payment, 2),
        "interest":     round(interest, 2),
        "principal":    round(payment - interest, 2),
        "pct_interest": round((interest / payment) * 100, 1),
    }

first_payment_split(principal=480000, annual_rate=0.06, years=30)
```

## Excel

With loan amount in `A2`, annual rate in `B2`, and monthly payment (from `PMT()`) in `C2`:

First-month interest:
```swift
=A2*(B2/12)
```

First-month principal:
```swift
=C2-A2*(B2/12)
```
:::
