---
title: "Savings Strategies"
---

```{r}
#| label: setup-r
#| message: false
#| warning: false
#| include: false
source("_common.R")
```

```{python}
#| label: setup-py
#| include: false
```

## Book Topics

- **Pay yourself first**: automate savings before anything else hits the checking account
- **Savings rate**: the single most important financial metric; higher rate = faster wealth accumulation regardless of returns
- **Benchmarks**: <10% danger zone; 10–20% average; 20–30% strong; 30%+ path to early financial independence
- **Sinking funds**: pre-save for known future expenses (car insurance, vacation, gifts) so they don't hit as surprises
- **Target**: save as much as you can *sustain*; a consistent 15% beats a sporadic 30%

## Savings Rate

The savings rate tells you how much of your income you're keeping, expressed as a percentage.

**Formula:** Savings Rate = (Income − Spending) ÷ Income × 100

```{r}
#| label: savings-rate-bar-model
#| echo: false
#| fig-cap: "Annual dollars saved at four savings-rate tiers on a $60,000 income"
#| fig-asp: 0.50
library(ggplot2)
income <- 60000
tiers <- data.frame(
  label = c("8% (danger zone)", "15% (average)", "25% (strong)", "35% (FI path)"),
  rate  = c(0.08, 0.15, 0.25, 0.35),
  fill  = c("#f44242", "#0e9aa7", "#48a56a", "#86ddcd"),
  stringsAsFactors = FALSE
)
tiers$saved   <- income * tiers$rate
tiers$bar_lbl <- paste0("saves $", format(tiers$saved, big.mark = ",", scientific = FALSE), "/yr")
tiers$label   <- factor(tiers$label, levels = rev(tiers$label))

ggplot(tiers, aes(x = label, y = saved, fill = label)) +
  geom_col(width = 0.6, colour = "#1C1C1E", linewidth = 0.3) +
  geom_text(aes(label = bar_lbl), hjust = -0.08, fontface = "bold",
            size = 3.5, colour = "#1C1C1E") +
  coord_flip() +
  scale_fill_manual(values = setNames(tiers$fill, tiers$label)) +
  scale_y_continuous(labels = function(x) paste0("$", format(x, big.mark = ",")),
                     limits = c(0, income * 0.45)) +
  labs(x = NULL, y = "Annual savings (on $60,000 income)") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank())
```

::: panel-tabset
## R

```{r}
#| label: savings-rate-r
savings_rate <- function(income, spending) {
  ((income - spending) / income) * 100
}

savings_rate(income = 5000, spending = 3500)
```

## Python

```{python}
#| label: savings-rate-py
def savings_rate(income, spending):
    return ((income - spending) / income) * 100

savings_rate(income=5000, spending=3500)
```

## Excel

With income in `A2` and spending in `B2`:

```swift
=((A2-B2)/A2)*100
```
:::
